The evolution of gene expression in the term placenta of viviparous mammals

### 2016-03-07 manuscript/manuscript.Rnw
use Spalax instead of Nannospalax

% species names
\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
\newcommand{\galili}[0]{\textit{N. galili}\xspace}
\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}

@@

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Nannospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\caption{Species tree for organisms studied}
\label{fig:species_tree}
\end{figure}

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top
5\% of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology
using the 1:1 \human ortholog obtained

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
The top 50 expressed genes in each genome analyzed are shown in
\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

\begin{table}[h]
  \footnotesize
<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
output.top.genes.table(top.genes.table(species.name="nannospalax galili"))
@ 
\caption{\tableexplanation{\galili}{tab:all_nannospalax_galili_placenta}}
\label{tab:top50_n_galili_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \galili are shown in
\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_nannospalax_galili_placenta}.
Components of annexin complexes are highly expressed
(\Sexpr{gene.expression("nannospalax_galili","S100a6",complete=TRUE)},
\Sexpr{gene.expression("nannospalax_galili","S100a11",complete=TRUE)},
\Sexpr{gene.expression("nannospalax_galili","S100a10",complete=TRUE)},
and
\Sexpr{gene.expression("nannospalax_galili","Anxa2",complete=TRUE)}),
as are members of the thymosin family
(\Sexpr{gene.expression("nannospalax_galili","Tmsb4x",complete=TRUE)}
and
\Sexpr{gene.expression("nannospalax_galili","Tmsb10",complete=TRUE)}),
various cytoskeleton related proteins
(\Sexpr{gene.expression("nannospalax_galili","Actc1",complete=TRUE)},
\Sexpr{gene.expression("nannospalax_galili","Actg2",complete=TRUE)},
\Sexpr{gene.expression("nannospalax_galili","Acta2",complete=TRUE)},
\Sexpr{gene.expression("nannospalax_galili","Myl9",complete=TRUE)},
\Sexpr{gene.expression("nannospalax_galili","Vim",complete=TRUE)}), and
transmembrane protein of currently unknown function, 
\Sexpr{gene.expression("nannospalax_galili","Gm10134",complete=TRUE)}.

>>>>

% species names
\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
\newcommand{\galili}[0]{\textit{S. galili}\xspace}
\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}

@@

Although the morphology of the mammalian placenta has been well
characterized in a number of species
\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
molecular environment is only beginning to be understood from a
genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
Studies of certain model species (i.e., \textit{Homo sapiens},
\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
divergence in the genetics of the placenta, especially in later stages
of pregnancy
\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
For example, each lineage has evolved expansive placenta-specific gene
families with expression patterns that can vary in space and time
\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
These include independent expansions of the prolactins in \mouse,
\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
in \cow, genes related to the thymus-specific serine protease
\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
and chorionic gonadotropin beta genes in hominoid primates
\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
Expression patterns of specific genes also can vary widely between
species. For example, the expression of orthologous genes in the
placenta have been found to vary substantially between \human and
\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
recruitment of \textit{SIGLEC6} to placental expression in \human
differs from both \textit{Pan troglodytes} and \mouse
\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.

@@

\begin{table}
  \begin{tabular}{c c c c}
    \toprule
    Species                       & Gestation (days) & Litter Size  & Source \\
    \midrule
    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Spalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
    \bottomrule
  \end{tabular}
  \caption{Gestational periods in days and average liter sizes for studied species.}
  \label{tab:gestational_periods}
\end{table}

@@

\begin{figure}
<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
species.tree <- 
    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Spalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
pp <- ggtree(species.tree)+
    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
    scale_x_continuous(limits=c(0,10))
print(pp %>%
      ## monodelphis
      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
      ## Armadillo
      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
      ## elephant
      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
      ## dog
      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
      ## horse
      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
      ## pig
      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
      ## sheep
      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
      ## cow
      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
      ## s carmeli
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
      ## n galili
      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
      ## mouse
      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
      ## bonobo
      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
      ## spider monkey
      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
      ## human
      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
      )
@ 
\caption{Species tree for organisms studied}
\label{fig:species_tree}
\end{figure}

@@

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using topGO
\cite{topGO:2.20.0}. Gene Ontology clusters were considered
significantly enriched if the \ac{FDR} as estimated by the Benjamini
and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
$p$ values obtained from the Fisher Exact Test were less than or equal
to 0.05. Enrichment tests were considered with two gene lists per
transcriptome: annotated genes with expression levels of the top
5\% of the frequency distribution of \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
the top 100 genes with the highest \ac{FPKM} (Supplemental
\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
Due to the paucity of annotated gene ontology terms for some species,
gene lists were compared using the \human Gene Ontology
using the 1:1 \human ortholog obtained

@@

Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
alignment statistics for all species. For transcriptomes sequenced
with 75 bp %
%
(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
%
For transcriptomes sequenced with 150 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
%
%
For transcriptomes sequenced with 35 bp, %
% The mismatch here is because we're reporting trimmed reads here, not
% full length reads
(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
%
,
the number of trimmed reads ranged from %
\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
%
to %
\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
%
with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
of reads mapping to the
respective reference genome (\cref{tab:ref_genome_aligned_species}).
The top 50 expressed genes in each genome analyzed are shown in
\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

@@

\begin{table}[h]
  \footnotesize
<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
output.top.genes.table(top.genes.table(species.name="spalax galili"))
@ 
\caption{\tableexplanation{\galili}{tab:all_spalax_galili_placenta}}
\label{tab:top50_n_galili_placenta}
\end{table}

The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
in the term placenta of \galili are shown in
\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
genes in Supplemental \cref{tab:all_spalax_galili_placenta}.
Components of annexin complexes are highly expressed
(\Sexpr{gene.expression("spalax_galili","S100a6",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","S100a11",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","S100a10",complete=TRUE)},
and
\Sexpr{gene.expression("spalax_galili","Anxa2",complete=TRUE)}),
as are members of the thymosin family
(\Sexpr{gene.expression("spalax_galili","Tmsb4x",complete=TRUE)}
and
\Sexpr{gene.expression("spalax_galili","Tmsb10",complete=TRUE)}),
various cytoskeleton related proteins
(\Sexpr{gene.expression("spalax_galili","Actc1",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","Actg2",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","Acta2",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","Myl9",complete=TRUE)},
\Sexpr{gene.expression("spalax_galili","Vim",complete=TRUE)}), and
transmembrane protein of currently unknown function, 
\Sexpr{gene.expression("spalax_galili","Gm10134",complete=TRUE)}.
